Emulation of high multiplicity matrix elements for electron-positron and hadron-hadron colliders

### 2022-11-17 mc.tex
changes to chapter 3

\documentclass[main.tex]{subfiles}
\begin{document}
\chapter{Monte Carlo Event Generators}\label{chap:MC}
    At particle collider experiments,
    the collisions between incoming particles can lead
    to highly complex final-states with a large number
    of particles. In a hadron collider, such as the LHC,
    the particles we observe are not the constituents
    of the proton, the quarks and gluons. Instead we observe
    hadrons which are the bound states of these partons
    due to the property of confinement at low energy scales.

    In order to describe the entire process
    from the initial energetic collision to the production
    of the lower energy hadrons, it is customary to
    split up the process into stages characterised by
    their kinematic hardness (c.f. Section~\ref{sec:factorisation}).
    The standard tools used
    to simulate these stages of an event are general-purpose
    Monte Carlo event generators (event generators), see
    for example Reference \cite{Buckley:2011ms} for a review.
    These are indispensable tools that help
    bridge the gap between theoretical predictions
    and experimental measurements with nearly all
    analyses made at the ATLAS and CMS experiments
    involving some form of event generator usage.
    General-purpose event generators
    such as Herwig \cite{Bahr:2008pv,Bellm:2015jjp},
    Pythia \cite{Sjostrand:2006za,Bierlich:2022pfr},
    and Sherpa \cite{Gleisberg:2008ta,Sherpa:2019gpd} all broadly share 
    the same features at a high-level, but each framework
    carries subtle differences. Hence, it is common
    to use many different event generators
    in one analysis to obtain an estimate of the uncertainty
    due to the different approaches in each framework.

@@

    \section{Monte Carlo integration}\label{sec:MC_integration}
    As seen in Section~\ref{sec:xs}, the task of
    computing partonic cross-sections boils down
    to integrating the matrix elements over the
    final-state phase-space. Whilst it is possible to
    carry out these integrals analytically for the simplest
    cases, the integral quickly becomes intractable when the
    final-state multiplicity becomes large. For an $n$-body
    final-state the dimensionality of the integral scales as $d=3n-4$,
    where momentum conservation and on-shell conditions
    have already been applied to reduce the dimensionality.
    Therefore, instead of carrying out the integrals analytically,
    numerical methods become more attractive. The current
    prevailing method is Monte Carlo integration which
    has no scaling with the dimensionality of the integral,
    and so can be used for arbitrarily high multiplicity final-states
    with no performance penalty compared to lower multiplicities.
    The purpose of this chapter is to give a brief overview
    of Monte Carlo integration in the context of event
    generation, a more in-depth treatment of the
    specifics of Monte Carlo methods can be found in, for example, Reference
    \cite{James:1980yn}.

    The basic idea of Monte Carlo integration is to approximate
    the integral of a function $f(\vec{x})$ by sampling it
    in $d$-dimensional parameter space points $\vec{x}$
    \begin{equation}\label{eqn:MC_integration}
        I = \int_{V} \mathrm{d}^{d}x \; f(\vec{x}) \approx \langle I(f) \rangle_{x} = V \dfrac{1}{N} \sum_{i=1}^{N} f(\vec{x}_{i}) = V \langle f \rangle_{x} \, ,
    \end{equation}
    where the points $\vec{x}$ are randomly,
    uniformly sampled in the volume $V$. $\langle f \rangle_{x}$
    denotes the average of $f$ over these uniformly sampled
    points. The law of large numbers ensures that the
    estimator $\langle I(f) \rangle_{x}$ will converge to
    the true value $I$ for $N \rightarrow \infty$. As a
    result of the central limit theorem, the function evaluations
    $f(\vec{x}_{i})$ will follow a Gaussian distribution with
    mean $\langle I(f) \rangle_{x}$,
    meaning an error estimate on the estimator is given by
    the standard error of the mean
    \begin{equation}\label{eqn:MC_error}
        \sigma_{I} = V \sqrt{\dfrac{\langle f^{2} \rangle_{x} - \langle f \rangle_{x}^{2}}{N}}
    \end{equation}
    which falls as $1/\sqrt{N}$, with no dependence
    on $d$.

    An algorithm that samples points $\vec{x}$ is called
    an integrator. To make use of the random numbers sampled,
    they need to be mapped to quantities that are useful in a simulation
    setting. The most commonly used mapping is to map
    the random numbers to four-momenta. An example of a
    simple integrator is the RAMBO algorithm \cite{Kleiss:1985gy,Platzer:2013esa}
    which samples four-momenta space uniformly. Since the integrator
    now effectively samples in momentum space, we refer to the points
    generated as phase-space points and so the terms integrator
    and phase-space generator are used interchangeably.

@@

    One such method is importance sampling which
    aims to reduce the variance via a remapping of the
    uniformly sampled variables $\vec{x}$
    to a more suitable non-uniform distribution.
    In the context of particle physics,
    this corresponds to using a mapping of random numbers
    to four-momenta that captures the distribution of the
    matrix elements and their divergences, i.e. samples
    more frequently in regions where the matrix element
    is large in magnitude. Integrators inspired by the pole
    structure of multi-parton QCD processes have been
    studied in References \cite{Draggiotis:2000gm,vanHameren:2002tc}.

    Another common variance reduction technique is
    stratified sampling which divides the integration
    volume into sub-volumes, or bins. The overall integral
    and variances are the sums of the partial results
    in each bin. The overall variance is minimised when
    the contributions from each bin are equal. This
    is achieved by sampling more from bins where the integral
    is rapidly fluctuating (and not necessarily large),
    and less from bins where the integral has less fluctuations.

    In practice, state-of-the-art event generators
    combine the concepts of importance sampling and stratified
    sampling with adaptive sampling \cite{Lepage:1977sw,Lepage:2020tgj}
    and multi-channelling methods \cite{Kleiss:1994qy,Ohl:1998jn}
    to achieve faster rates of convergence.

    \section{Event unweighting}\label{sec:unweighting}
    Once a suitable configuration of momenta
    has been found by the integrator, this point forms the basis of an
    event. The matrix element of interest can be evaluated at
    this phase-space point at the desired order in perturbation
    theory (see Section~\ref{sec:me_generators}). Additionally, there will be an associated Jacobian
    from the mapping from random variables to the four-momenta.
    This Jacobian, along with the PDF weights and flux factors,
    are referred to as the phase-space weight, $J_{i}$. Taking
    this phase-space weight together
    with the matrix element forms the event weight\footnote{For
    the more sophisticated sampling methods employed in event generators
    there will be additional weights involved.}
    \begin{equation}\label{eqn:event_weight}
        w_{i} = |\mathcal{M}|_{i}^{2} J_{i} \, .
    \end{equation}
    The event weight can be interpreted as the probability
    of the event occurring at an experiment \cite{Plehn:2009nd}.
    By generating a number of events in this manner,
    and histogramming them, a sample of weighted events is created.

@@

    \begin{algorithm}
        \caption{Rejection sampling for unweighting events}\label{alg:unweighting}
        \While{unweighting}{
            generate random phase-space point $\vec{x}$\;
            evaluate event weight $w \gets w(\vec{x})$\;
            generate uniform random number $r \gets$ Random(0, 1)\;
            \If{$w / w_{\mathrm{max}} > r$}{
                \KwRet $\vec{x}$ and $\tilde{w} \gets 1$\;
                % $\tilde{w} \gets \max(1,\; w / w_{\mathrm{max}})$\;
            }
        }
    \end{algorithm}

@@

    An alternative to improving the unweighting efficiency
    itself is to accelerate the process of unweighting itself. This
    was explored in Reference \cite{Danziger:2021eeg} wherein a
    novel two-stage unweighting procedure employing a surrogate
    model of the exact event weight was used to decrease the amount of
    time spent unweighting events. This idea will be elaborated on
    in Chapter~\ref{chapter:fame3} where we build upon
    Reference \cite{Danziger:2021eeg} by increasing the fidelity of
    the surrogate model.

    \section{Anatomy of an event}\label{sec:anatomy_event}
    The simulation of an event can be split up into three
    main stages: the hard scattering process, followed by
    the parton shower, and finally hadronisation.
    This is motivated by the factorisation theorem (Section~\ref{sec:factorisation})
    which separates the high energy regimes from the low
    energy regimes.

@@

    As seen in Eq. (\ref{eqn:matrix_element}),
    the exact matrix element is approximated by a perturbative
    series in a coupling constant. By now it is standard to
    have at least the LO and NLO contributions in this expansion.
    The calculations of
    tree-level \cite{Mangano:2002ea,Krauss:2001iv,Gleisberg:2008fv,Cafarella:2007pc,Alwall:2011uj,Kilian:2007gr}
    and one-loop \cite{Berger:2008sj,Cullen:2011ac,Cullen:2014yla,Bevilacqua:2011xh,Hirschi:2011pa,Alwall:2014hca,Frederix:2018nkq,Campbell:1999ah,Campbell:2021vlt,Badger:2012pg,Cascioli:2011va,Buccioni:2019sur,Actis:2016mpe,Denner:2017wsf}
    matrix elements are now largely fully automated
    in NLO QCD and electroweak corrections.
    Many of the modern matrix element generators are supplied
    with a common generic interface to event generators,
    the Binoth Les Houches Accord \cite{Binoth:2010xt,Alioli:2013nda}.
    For more information on the machinery developed to
    compute matrix elements at tree-level and
    one-loop level required for SM predictions, see for instance Reference \cite{Campbell:2017hsr}.
    NNLO and N$^{3}$LO QCD corrections have been computed
    for select processes but have yet to be reach the
    level of automation of NLO.
    For a review of recent advancements see References \cite{Caola:2022ayt,Campbell:2022qmc}.

@@

    \subsection*{Hadronisation}\label{sec:hadronisation}
    The simulation of the formation of hadrons in the low-energy regime
    of an event generator follow phenomenological, non-perturbative models.
    These hadronisation models require the combination
    of quarks and antiquarks to form colourless hadrons.
    The two main classes of models used to simulate
    hadronisation are the cluster models \cite{Webber:1983if,Winter:2003tt}
    and string models \cite{Andersson:1983ia}.
    For more discussion on hadronisation
    see for example Reference \cite{Webber:1999ui}.

@@

    The most widely adopted approach to estimate the
    theoretical uncertainties associated with these
    missing higher order terms is to carry out a seven-point
    scale variation. Typically in a calculation,
    $\mu_{0} = \mu_{F} = \mu_{R}$ where $\mu_{0}$ depends on the specifics
    of the process of interest. Taking the value of the
    partonic cross-section at this scale as the central value,
    the seven variations correspond to multiplying ($\mu_{F}, \, \mu_{R}$) by the factors 
    \begin{equation}\label{eqn:scale_variations}
        S \in \left\{(1, 1), (\sfrac{1}{2}, \sfrac{1}{2}), (2, 2), (\sfrac{1}{2}, 1), (1, \sfrac{1}{2}), (2, 1), (1, 2) \right\} \, ,
    \end{equation}
    and re-evaluating the cross-section at these scales.
    The deviations from the central cross-section value
    forms an estimate of the uncertainty in the perturbatively
    computed cross-section, when compared to the true cross-section.
    This can be written as
    \begin{equation}
        \hat{\sigma}(\mu_{F}, \mu_{R}) = \hat{\sigma}(\mu_{0}, \mu_{0}) \pm \delta \, ,
    \end{equation}
    where $\delta$ is determined from the scale variations.
    However, there is no single, well agreed upon definition
    of $\delta$ \cite{Cacciari:2011ze}.
    The reasoning behind this approach is that $\delta$
    is approximately the same order as the true missing
    higher order terms.

>>>>

\documentclass[main.tex]{subfiles}
\begin{document}
\chapter{Monte Carlo Event Generators}\label{chapter:MC}
    At particle collider experiments,
    the collisions between incoming particles can lead
    to highly complex final-states with a large number
    of particles. In a hadron collider, such as the LHC,
    the particles we observe are not the constituents
    of the proton, the quarks and gluons. Instead we observe
    hadrons which are the bound states of these partons
    due to the property of confinement at low energy scales.

    In order to describe the entire process
    from the initial energetic collision to the production
    of the lower energy hadrons, it is customary to
    split up the process into stages characterised by
    their kinematic hardness (c.f. Section~\ref{sec:factorisation}).
    The standard tools used
    to simulate these stages of an event are general-purpose
    Monte Carlo event generators, see
    Ref. \cite{Buckley:2011ms} for a review.
    These are indispensable tools that help
    bridge the gap between theoretical predictions
    and experimental measurements with nearly all
    analyses made at the ATLAS and CMS experiments
    involving some form of event generator usage.
    General-purpose event generators
    such as {\Herwig} \cite{Bahr:2008pv,Bellm:2015jjp},
    {\Pythia} \cite{Sjostrand:2006za,Bierlich:2022pfr},
    and {\Sherpa} \cite{Gleisberg:2008ta,Sherpa:2019gpd} all broadly share 
    the same features at a high-level, but each framework
    carries subtle differences. Hence, it is common
    to use many different event generators
    in one analysis to obtain an estimate of the uncertainty
    due to the different approaches in each framework.

@@

    \section{Monte Carlo integration}\label{sec:MC_integration}
    As seen in Section~\ref{sec:xs}, the task of
    computing partonic cross-sections boils down
    to integrating the matrix elements over the
    final-state phase-space. Whilst it is possible to
    carry out these integrals analytically for the simplest
    cases, the integral quickly becomes intractable when the
    final-state multiplicity becomes large. For an $n$-body
    final-state the dimensionality of the integral scales as $d=3n-4$,
    where momentum conservation and on-shell conditions
    have already been applied to reduce the dimensionality.
    Therefore, instead of carrying out the integrals analytically,
    numerical methods become more attractive. The current
    prevailing method is Monte Carlo integration which
    has no scaling with the dimensionality of the integral,
    and so can be used for arbitrarily high multiplicity final-states
    with no performance penalty compared to lower multiplicities.
    The purpose of this chapter is to give a brief overview
    of Monte Carlo integration in the context of event
    generation, a more in-depth treatment of the
    specifics of Monte Carlo methods can be found in Ref.
    \cite{James:1980yn}.

    The basic idea of Monte Carlo integration is to approximate
    the integral of a function $f(\mathbf{x})$ by sampling it
    in $d$-dimensional parameter space points $\mathbf{x}$
    \begin{equation}\label{eqn:MC_integration}
        I = \int_{V} \mathrm{d}^{d}x \; f(\mathbf{x}) \approx \langle I(f) \rangle_{x} = V \dfrac{1}{N} \sum_{i=1}^{N} f(\mathbf{x}_{i}) = V \langle f \rangle_{x} \, ,
    \end{equation}
    where the points $\mathbf{x}$ are randomly,
    uniformly sampled in the volume $V$. $\langle f \rangle_{x}$
    denotes the average of $f$ over these uniformly sampled
    points. The law of large numbers ensures that the
    estimator $\langle I(f) \rangle_{x}$ will converge to
    the true value $I$ for $N \rightarrow \infty$. As a
    result of the central limit theorem, the function evaluations
    $f(\mathbf{x}_{i})$ will follow a Gaussian distribution with
    mean $\langle I(f) \rangle_{x}$,
    meaning an error estimate on the estimator is given by
    the standard error of the mean
    \begin{equation}\label{eqn:MC_error}
        \sigma_{I} = V \sqrt{\dfrac{\langle f^{2} \rangle_{x} - \langle f \rangle_{x}^{2}}{N}}
    \end{equation}
    which falls as $1/\sqrt{N}$, with no dependence
    on $d$.

    An algorithm that samples points $\mathbf{x}$ is called
    an integrator. To make use of the random numbers sampled,
    they need to be mapped to quantities that are useful in a simulation
    setting. The most commonly used mapping is to map
    the random numbers to four-momenta. An example of a
    simple integrator is the RAMBO algorithm \cite{Kleiss:1985gy,Platzer:2013esa}
    which samples four-momenta space uniformly. Since the integrator
    now effectively samples in momentum space, we refer to the points
    generated as phase-space points and so the terms integrator
    and phase-space generator are used interchangeably.

@@

    One such method is importance sampling which
    aims to reduce the variance via a remapping of the
    uniformly sampled variables $\mathbf{x}$
    to a more suitable non-uniform distribution.
    In the context of particle physics,
    this corresponds to using a mapping of random numbers
    to four-momenta that captures the distribution of the
    matrix elements and their divergences, i.e. samples
    more frequently in regions where the matrix element
    is large in magnitude. Integrators inspired by the pole
    structure of multi-parton QCD processes have been
    studied in Refs. \cite{Draggiotis:2000gm,vanHameren:2002tc}.

    Another common variance reduction technique is
    stratified sampling which divides the integration
    volume into sub-volumes, or bins. The overall integral
    and variance are the sums of the partial results
    in each bin. The overall variance is minimised when
    the contributions from each bin are equal. This
    is achieved by sampling more from bins where the integral
    is rapidly fluctuating (and not necessarily large),
    and less from bins where the integral has fewer fluctuations.

    In practice, state of the art event generators
    combine the concepts of importance sampling and stratified
    sampling with adaptive sampling \cite{Lepage:1977sw,Lepage:2020tgj}
    and multi-channelling methods \cite{Kleiss:1994qy,Ohl:1998jn}
    to achieve faster rates of convergence.

    \section{Event unweighting}\label{sec:unweighting}
    Once a suitable configuration of momenta
    has been found by the integrator, this phase-space point forms the basis of an
    event. The matrix element of interest can be evaluated at
    this phase-space point at the desired order in perturbation
    theory (see Section~\ref{sec:me_generators}).
    Additionally, there will be an associated Jacobian
    from the mapping from random variables to the four-momenta.
    This Jacobian, along with the PDF weights and flux factors,
    are referred to as the phase-space weight, $J_{i}$. Taking
    this phase-space weight together
    with the matrix element forms the event weight\footnote{For
    the more sophisticated sampling methods employed in event generators
    there will be additional weights involved.}
    \begin{equation}\label{eqn:event_weight}
        w_{i} = |\mathcal{M}|_{i}^{2} J_{i} \, .
    \end{equation}
    The event weight can be interpreted as the probability
    of the event occurring at an experiment \cite{Plehn:2009nd}.
    By generating a number of events in this manner,
    and histogramming them, a sample of weighted events is created.

@@

    \begin{algorithm}
        \caption{Rejection sampling for unweighting events}\label{alg:unweighting}
        \While{unweighting}{
            generate random phase-space point $\mathbf{x}$\;
            evaluate event weight $w \gets w(\mathbf{x})$\;
            generate uniform random number $r \gets$ Random(0, 1)\;
            \If{$w / w_{\mathrm{max}} > r$}{
                \KwRet $\mathbf{x}$ and $\tilde{w} \gets 1$\;
                % $\tilde{w} \gets \max(1,\; w / w_{\mathrm{max}})$\;
            }
        }
    \end{algorithm}

@@

    An alternative to improving the unweighting efficiency
    itself is to accelerate the process of unweighting itself. This
    was explored in Ref. \cite{Danziger:2021eeg} wherein a
    novel two-stage unweighting procedure employing a surrogate
    model of the exact event weight was used to decrease the amount of
    time spent unweighting events. This idea will be elaborated on
    in Chapter~\ref{chapter:fame3} where we build upon
    Ref. \cite{Danziger:2021eeg} by increasing the fidelity of
    the surrogate model.

    \section{Anatomy of an event}\label{sec:anatomy_event}
    The simulation of an event can be split up into three
    main stages: the hard scattering process, followed by
    the parton shower, and finally hadronisation.
    This is motivated by the factorisation theorem (c.f. Section~\ref{sec:factorisation})
    which separates the high energy regimes from the low
    energy regimes.

@@

    As seen in Eq. (\ref{eqn:matrix_element}),
    the exact matrix element is approximated by a perturbative
    series in a coupling parameter. By now it is standard to
    have at least the LO and NLO contributions in this expansion.
    The calculations of
    tree-level \cite{Mangano:2002ea,Krauss:2001iv,Gleisberg:2008fv,Cafarella:2007pc,Alwall:2011uj,Kilian:2007gr}
    and one-loop \cite{Berger:2008sj,Cullen:2011ac,Cullen:2014yla,Bevilacqua:2011xh,Hirschi:2011pa,Alwall:2014hca,Frederix:2018nkq,Campbell:1999ah,Campbell:2021vlt,Badger:2012pg,Cascioli:2011va,Buccioni:2019sur,Actis:2016mpe,Denner:2017wsf}
    matrix elements are now largely fully automated
    in NLO QCD and electroweak corrections.
    Many of the modern matrix element generators are supplied
    with a common generic interface to event generators,
    the Binoth Les Houches Accord \cite{Binoth:2010xt,Alioli:2013nda}.
    For more information on the machinery developed to
    compute matrix elements at tree-level and
    one-loop level required for SM predictions, see Ref. \cite{Campbell:2017hsr}.
    NNLO and N$^{3}$LO QCD corrections have been computed
    for select processes but have yet to be reach the
    level of automation of NLO.
    For a review of recent advancements see Refs. \cite{Caola:2022ayt,Campbell:2022qmc}.

@@

    \subsection*{Hadronisation}\label{sec:hadronisation}
    The simulation of the formation of hadrons in the low-energy regime
    of an event generator follows phenomenological, non-perturbative models.
    These hadronisation models require the combination
    of quarks and antiquarks to form colourless hadrons.
    The two main classes of models used to simulate
    hadronisation are the cluster models \cite{Webber:1983if,Winter:2003tt}
    and string models \cite{Andersson:1983ia}.
    For more discussion on hadronisation
    see Ref. \cite{Webber:1999ui}.

@@

    The most widely adopted approach to estimate the
    theoretical uncertainties associated with these
    missing higher order terms is to carry out a seven point
    scale variation. Typically in a calculation,
    $\mu_{0} = \mu_{F} = \mu_{R}$ where $\mu_{0}$ depends on the specifics
    of the process of interest. Taking the value of the
    partonic cross-section at this scale as the central value,
    the seven variations correspond to multiplying ($\mu_{F}, \, \mu_{R}$) by the factors 
    \begin{equation}\label{eqn:scale_variations}
        S \in \left\{(1, 1), (\sfrac{1}{2}, \sfrac{1}{2}), (2, 2), (\sfrac{1}{2}, 1), (1, \sfrac{1}{2}), (2, 1), (1, 2) \right\} \, ,
    \end{equation}
    and re-evaluating the cross-section at these scales.
    The deviations from the central cross-section value
    forms an estimate of the uncertainty in the perturbatively
    computed cross-section, when compared to the true cross-section.
    This can be written as
    \begin{equation}
        \hat{\sigma}(\mu_{F}, \mu_{R}) = \hat{\sigma}(\mu_{0}, \mu_{0}) \pm \delta \, ,
    \end{equation}
    where $\delta$ is determined from the scale variations.
    However, there is no single, well agreed upon definition
    of $\delta$ \cite{Cacciari:2011ze}.
    The rationale behind taking this approach is that $\delta$
    is approximately the same order as the true missing
    higher order terms.
